Responsible Machine Learning Techniques\\\large{Interpretable Models, Post-hoc Explanation, and Discrimination Testing}

### 2020-01-27 article-information-2019.tex
initial chop, -4 pages

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Notation}\label{ssec:not}

@@

The US HMDA law, originally enacted in 1975, requires many financial institutions that originate mortgage products to provide certain data about many of the mortgage-related products that they either deny or originate on an annual basis. This information is first provided to the Consumer Financial Protection Bureau (CFPB), which subsequently releases some of the data to the public. Regulators often use HMDA data to, ``...show whether lenders are serving the housing needs of their communities; they give public officials information that helps them make decisions and policies; and they shed light on lending patterns that could be discriminatory.''\textsuperscript{5} In addition to regulatory use, public advocacy groups use these data for similar purposes, and the lenders themselves use the data to benchmark their community outreach relative to their peers.  The publicly available data that the CFPB releases includes information such as the lender, the type of loan, loan amount, loan to value (LTV) ratio, debt to income (DTI) ratio, and other important financial descriptors. The data also include information on each borrower and co-borrower’s race, ethnicity, gender, and age. Because the data includes information on these protected class characteristics, certain metrics that can be indicative of discrimination in lending can be calculated directly using the HDMA data.

The mortgage dataset analyzed here is a random sample of consumer-anonymized loans from the HDMA database. These loans are a subset of all originated mortgage loans in the 2018 HMDA data that were chosen to represent a relatively comparable group of consumer mortgages.  A selection of features is used to predict whether a loan is \textit{high-priced}, i.e., the annual percentage rate (APR) charged was 150 basis points (1.5\%) or more above a survey-based estimate of other similar loans offered around the time of the given loan.  After data cleaning and preprocessing to encode categorical features and create missing markers, the mortgage data contains ten input features and the binary outcome, \textit{high-priced}. The data is split into a training set with 160,338 loans and a marker for 5 fold cross validation and a test set containing 39,662 loans. While lenders would almost certainly use more information than the selected features to determine whether to offer and originate a high-priced loan, the selected input features (LTV ratio, DTI ratio, property value, loan amount, introductory interest rate, customer income, etc.) are likely to be some of the most influential factors that a lender would consider.  Ultimately, the HMDA data represent the most comprehensive source of data on highly-regulated mortgage lending that is publicly available, which makes it an ideal dataset to use for the types of analyses set forth in Sections \ref{sec:m_and_m} and \ref{sec:res}.

@@

Herein, two $g^\text{MGBM}$ models are trained.  One on the simulated data and one on the mortgage data. In both cases, positive and negative monotonic constraints for each $X_j$ are selected using domain knowledge, random grid search is used to determine other hyperparameters, and five-fold cross validation and test partitions are used for model assessment. For exact parameterization of the two $g^\text{MGBM}$ models, see the software resources referenced in Subsection \ref{ssec:soft}.

@@

Here, each $g^\text{XNN}$ is trained by mini-batch stochastic gradient descent (SGD) on the simulated data and mortgage data. Each $g^\text{XNN}$ is assessed in five training folds and in a test data partition. $L_1$ regularization is applied to both the projection and combination layers to induce a sparse and interpretable model, where each $n_k$ subnetwork and corresponding combination layer $\gamma_k$ are ideally associated with an important $X_j$ or combination thereof. The $g^\text{XNN}$ models appear highly sensitive to weight initialization and batch size. Be aware that $g^{XNN}$ model architectures may require manual and judicious feature selection due to long training times. For more details regarding $g^\text{XNN}$ training, see the software resources in Subsection \ref{ssec:soft}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Partial Dependence and Individual Conditional Expectation}\label{ssec:pd_ice}

PD plots are a widely-used method for describing and plotting the average predictions of a complex model $g$ across some partition of data $\mathbf{X}$ for some interesting input feature $X_j$ \cite{esl}. ICE plots are a newer method that describes the local behavior of $g$ for a single instance $\mathbf{x} \in \mathcal{X}$ \cite{ice_plots}. PD and ICE can be overlaid in the same plot to compensate for known weaknesses of PD (e.g., inaccuracy in the presence of strong interactions and correlations \cite{ice_plots}, \cite{ale_plot}), to identify interactions modeled by $g$, and to create a holistic global and local portrait of the predictions for some $g$ and $X_j$ \cite{ice_plots}.

@@

Shapley explanations are a class of additive, locally accurate feature contribution measures with long-standing theoretical support \cite{shapley}, \cite{shapley1988shapley}. Shapley explanations are the only possible locally accurate and globally consistent feature contribution values, meaning that Shapley explanation values for input features always sum to $g(\mathbf{x})$ for some $\mathbf{x} \in \mathcal{X}$ and that Shapley explanation values should never decrease in magnitude for some $x_j$ when $g$ is changed such that $x_j$ truly makes a stronger contribution to $g(\mathbf{x})$ \cite{shapley}, \cite{tree_shap}. For some instance $\mathbf{x} \in \mathcal{X}$, Shapley explanations take the form:

Shapley values can be estimated in different ways, many of which are intractable for datasets with large $\mathcal{P}$. Tree SHAP is a specific implementation of Shapley explanations that relies on traversing internal decision tree structures to efficiently estimate the contribution of each $x_j$ for some $g(\mathbf{x})$ \cite{tree_shap}. Tree SHAP (SHapley Additive exPlanations) has been implemented efficiently in popular gradient boosting libraries such as \href{http://docs.h2o.ai/h2o/latest-stable/h2o-py/docs/modeling.html#h2ogradientboostingestimator}{\texttt{h2o}}, \href{https://lightgbm.readthedocs.io/en/latest/Python-Intro.html}{\texttt{LightGBM}}, and \href{https://xgboost.readthedocs.io/en/latest/python/python_intro.html}{\texttt{XGBoost}}, and Tree SHAP is used to calculate accurate and consistent global and local feature importance for MGBM models in Sections \ref{sec:m_and_m} and \ref{sec:res}. Deep SHAP is an approximate Shapley value technique that creates SHAP values for ANNs \cite{shapley}. Deep SHAP is implemented in the \href{https://github.com/slundberg/shap}{\texttt{shap}} package and is used to generate SHAP values for the two $g^{XNN}$ models discussed in Sections \ref{sec:m_and_m} and \ref{sec:res}.

@@

Because many technical and academic discussions of fairness in ML have been inconclusive\footnote{See: \href{https://www.youtube.com/watch?v=jIXIuYdnyyk}{Tutorial: 21 Fairness Definitions and Their Politics}.}, this text will draw on regulatory and legal standards that have been used for years in industries like Financial Services. The discussed metrics are also representative of fair lending analyses and pair well with the mortgage data. Before discussing the techniques, it is important to explain and draw a distinction between the two major types of discrimination recognized in US legal and regulatory settings, disparate treatment (DT), and disparate impact (DI). DT (which is loosely referred to as \textit{intentional discrimination}) occurs most often in an algorithmic setting when a model explicitly uses protected class status (e.g., race, sex) as an input feature or uses a feature that is so similar to protected class status that it essentially proxies for class membership.  With some limited exceptions, the use of these factors in an algorithm is illegal under several statutes in the US.\textsuperscript{4} DI, colloquially known as \textit{unintentional discrimination}, occurs when some element of a decisioning process includes a \textit{facially neutral} factor (i.e, a reasonable and valid predictor of response) that results in a disproportionate share of a protected class receiving an unfavorable outcome.  In modeling, this is most typically driven by a statistically important feature that is distributed unevenly across classes, which causes more frequent unfavorable outcomes for the protected class.  However, other factors, such as hyperparameter or algorithm choices, can drive DI.  Crucially, legality hinges on whether changing the model, for example exchanging one feature for another or altering the hyperparameters of an algorithm, can lead to a similarly predictive model with lower disparate impact.

The analyses and metrics herein focus on several measures of disparate impact that are commonly used in US litigation and regulatory settings. One is known as marginal effects (ME). ME is simply the difference between the percent of the control group members receiving a favorable outcome and the percent of the protected class members receiving a favorable outcome.

@@

Tables \ref{tab:sim_n_acc} and \ref{tab:sim_disc} show the results of the disparity tests using the simulated data for two hypothetical sets of class-control groups.  Several measures of disparities are shown, with the SMDs calculated using the probabilities from $g^{\text{MGBM}}$ and  $g^{\text{XNN}}$, and the false positive rates (FPRs), false negative rates (FNRs), their ratios, MEs, and AIRs calculated using a binary outcome based on a cutoff of 0.6 (anyone with probabilities of 0.6 or greater receives the favorable outcome).\footnote{The selection of which cutoff to use in production is typically based on the model’s use case, rather than one based solely on the statistical properties of the predictions themselves.  For example, a model developer at a bank might build a credit model where the F1 score is maximized at a delinquency probability cutoff of 0.15.  For purposes of evaluating the quality of the model, she may review confusion matrix statistics (accuracy, recall, precision, etc.) using cutoffs based on the maximum F1 score. But, because of its risk tolerance and other factors, the bank itself might be willing to lend to anyone with a delinquency probability of 0.18 or lower, which would mean that anyone who is scored at 0.18 or lower would receive an offer of credit.  Because disparity analyses are concerned with how people are affected by the way the model is used, it is essential that any confusion matrix-based metrics of disparity be calculated on the in-production classification decisions, rather than the cutoffs that are not related to what those affected by the model will experience.} 
Since $g^{\text{MGBM}}$ and $g^{\text{XNN}}$ assume that a higher score is favorable (as might be the case if the model were predicting responses to marketing offers), one might consider the relative FNRs as a measure of the class-control disparities. Table \ref{tab:sim_disc} shows that protected group 1 has higher relative FNRs under $g^{\text{XNN}}$ (1.13 vs. 1.06).  However, in Table \ref{tab:sim_n_acc} the overall FNRs were lower for  $g^{\text{XNN}}$ (0.357 vs. 0.401). This illustrates a danger in considering relative class-control metrics in isolation when comparing across models: despite the $g^\text{MGBM}$ appearing to be a relatively fairer model, more protected group 1 members experience negative outcomes using $g^{\text{MGBM}}$. This is because FNR accuracy improves for both the protected group 1 and control group 1, but members of control group 1 benefit more than those in protected group 1.  Of course, the choice of which model is truly fairer is a policy question.

\begin{table}[H]
\caption{Group size, accuracy, FPR, and FNR for $g^\text{MGBM}$ and $g^\text{XNN}$ on the simulated test data.} 
\centering
\begin{tabular}{cccccc}
\toprule
\textbf{Class}	& $\mathbf{N}$	& \textbf{Model} 												& \textbf{Accuracy}  								& \textbf{FPR}									& \textbf{FNR} \\
\midrule
Protected 1	& 3,057		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.770 \\ 0.771\end{tabular} 	& \begin{tabular}{@{}c@{}} 0.150 \\ 0.168\end{tabular} 	& \begin{tabular}{@{}c@{}} 0.401 \\ 0.357\end{tabular} \\	
\midrule			 
Control 1		& 16,943		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.739 \\ 0.756\end{tabular}	& \begin{tabular}{@{}c@{}} 0.129 \\ 0.163\end{tabular}	& \begin{tabular}{@{}c@{}} 0.378 \\ 0.314\end{tabular} \\
\midrule
Protected 2	& 9,916		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.758 \\ 0.762\end{tabular}	& \begin{tabular}{@{}c@{}} 0.169 \\ 0.186\end{tabular}	& \begin{tabular}{@{}c@{}} 0.331 \\ 0.302\end{tabular} \\	
\midrule
Control 2		& 10,084		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.729 \\ 0.756\end{tabular}	& \begin{tabular}{@{}c@{}} 0.091 \\ 0.138\end{tabular}	& \begin{tabular}{@{}c@{}} 0.420 \\ 0.332\end{tabular} \\		
\bottomrule
\label{tab:sim_n_acc}
\end{tabular}
\end{table}

@@

\begin{table}[H]
\caption{AIR, ME, SMD, FPR ratio, and FNR ratio for $g^\text{MGBM}$ and $g^\text{XNN}$ on the simulated test data.}
\centering
\begin{tabular}{cccccccc}
\toprule
\textbf{Model}		& \textbf{Protected Class}										& \textbf{Control Class}								& \textbf{AIR}									& \textbf{ME}									& \textbf{SMD}   								& \textbf{FPR Ratio}							& \textbf{FNR Ratio} \\
\midrule
$g^\text{MGBM}$	& \begin{tabular}{@{}c@{}} Protected 1 \\ Protected 2\end{tabular}		& \begin{tabular}{@{}c@{}} Control 1 \\ Control 2\end{tabular}	& \begin{tabular}{@{}c@{}} 0.752 \\ 1.10\end{tabular}	& \begin{tabular}{@{}c@{}} 9.7\% \\ -3.6\%\end{tabular}	& \begin{tabular}{@{}c@{}} -0.206 \\ 0.106\end{tabular}	& \begin{tabular}{@{}c@{}} 1.16 \\ 1.86\end{tabular}	& \begin{tabular}{@{}c@{}} 1.06 \\ 0.788\end{tabular}\\
\midrule 
$g^\text{XNN}$		& \begin{tabular}{@{}c@{}} Protected 1 \\ Protected 2\end{tabular}		& \begin{tabular}{@{}c@{}} Control 1 \\ Control 2\end{tabular}	& \begin{tabular}{@{}c@{}} 0.727 \\ 0.976\end{tabular}	& \begin{tabular}{@{}c@{}} 12.0\%\\ 1.0\%\end{tabular}	& \begin{tabular}{@{}c@{}} -0.274 \\ 0.001\end{tabular}	& \begin{tabular}{@{}c@{}} 1.03 \\ 1.35\end{tabular}	& \begin{tabular}{@{}c@{}} 1.13 \\ 0.907\end{tabular} \\										
\bottomrule
\label{tab:sim_disc}
\end{tabular}
\end{table}

@@

\begin{table}[H]
\caption{Group size, accuracy, FPR, and FNR for $g^\text{MGBM}$ and $g^\text{XNN}$ on the mortgage test data.}
\centering
\begin{tabular}{ccccccc}
\toprule
\textbf{Class}	& $\mathbf{N}$	& \textbf{Model} 												& \textbf{Accuracy} 									& \textbf{FPR}											& \textbf{FNR} \\
\midrule
Black 		& 2,608		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.654 \\ 0.702\end{tabular} 		& \begin{tabular}{@{}c@{}} 0.315 \\ 0.295\end{tabular} 			& \begin{tabular}{@{}c@{}} 0.457 \\ 0.308\end{tabular} \\	
\midrule			 
White 		& 28,361		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.817 \\ 0.857\end{tabular} 		& \begin{tabular}{@{}c@{}} 0.150 \\ 0.120\end{tabular}			& \begin{tabular}{@{}c@{}} 0.508 \\ 0.360\end{tabular}\\		
\midrule
Female 		& 8,301		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}}  0.768 \\ 0.822\end{tabular}		& \begin{tabular}{@{}c@{}} 0.208 \\ 0.158\end{tabular}			& \begin{tabular}{@{}c@{}} 0.402 \\ 0.322\end{tabular}\\
\midrule
Male 		& 13,166		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.785 \\	0.847\end{tabular} 		& \begin{tabular}{@{}c@{}} 0.182 \\ 0.131\end{tabular}			& \begin{tabular}{@{}c@{}} 0.497 \\ 0.347\end{tabular}\\
\bottomrule
\label{tab:mort_n_acc}
\end{tabular}
\end{table}

@@

\begin{table}[H]
\caption{AIR, ME, SMD, FPR ratio, and FNR ratio for $g^\text{MGBM}$ and $g^\text{XNN}$ on the mortgage test data.}
\centering
\begin{tabular}{cccccccc}
\toprule
\textbf{Model}	& \textbf{Protected Class}	& \textbf{Control Class}	& \textbf{AIR}	& \textbf{ME}	& \textbf{SMD}  & \textbf{FPR Ratio}		& \textbf{FNR Ratio} \\
\midrule
$g^\text{MGBM}$	& \begin{tabular}{@{}c@{}} Black \\ Female\end{tabular}		& \begin{tabular}{@{}c@{}} White \\ Male\end{tabular}	& \begin{tabular}{@{}c@{}} 0.776 \\ 0.948\end{tabular}	& \begin{tabular}{@{}c@{}} 18.3\% \\ 4.1\%\end{tabular}	& \begin{tabular}{@{}c@{}} 0.628 \\ 0.084\end{tabular}	& \begin{tabular}{@{}c@{}} 2.10 \\ 1.15\end{tabular}	& \begin{tabular}{@{}c@{}} 0.900 \\ 0.810\end{tabular} \\
\midrule 
$g^\text{XNN}$		& \begin{tabular}{@{}c@{}} Black \\ Female\end{tabular}		& \begin{tabular}{@{}c@{}} White \\ Male\end{tabular}	& \begin{tabular}{@{}c@{}} 0.743 \\ 0.955\end{tabular}	& \begin{tabular}{@{}c@{}} 21.4\% \\ 3.6\%\end{tabular}	& \begin{tabular}{@{}c@{}} 0.621 \\ 0.105\end{tabular}	& \begin{tabular}{@{}c@{}} 2.45 \\ 1.21\end{tabular}	& \begin{tabular}{@{}c@{}} 0.855 \\ 0.927\end{tabular} \\							
\bottomrule
\label{tab:mort_disc}
\end{tabular}
\end{table}

@@

The heightened interest in exploring and fixing algorithmic bias has led to a significant body of work focused in using ML to diminish discrimination in model outcomes \cite{friedler2019comparative}. Broadly, these can be defined into two groups: more traditional methods that mitigate discrimination by searching across possible algorithmic and feature specifications, and more recently developed approaches that change the algorithms or input data themselves in order to mitigate disparities. Many approaches that have been developed in the last five to seven years focus on altering the algorithm itself, preprocessing the data, or post-processing the predictions in order to diminish class-control correlations or dependencies. IBM’s \href{http://aif360.mybluemix.net/}{\texttt{AIF360}} toolkit provides an open source Python package for many of the metrics and techniques that have been developed recently.

>>>>

\usepackage{caption}

@@

@@

The mortgage dataset analyzed here is a random sample of consumer-anonymized loans from the HDMA database. These loans are a subset of all originated mortgage loans in the 2018 HMDA data that were chosen to represent a relatively comparable group of consumer mortgages.  A selection of features is used to predict whether a loan is \textit{high-priced}, i.e., the annual percentage rate (APR) charged was 150 basis points (1.5\%) or more above a survey-based estimate of other similar loans offered around the time of the given loan.  After data cleaning and preprocessing to encode categorical features and create missing markers, the mortgage data contains ten input features and the binary outcome, \textit{high-priced}. The data is split into a training set with 160,338 loans and a marker for 5 fold cross validation and a test set containing 39,662 loans. While lenders would almost certainly use more information than the selected features to determine whether to offer and originate a high-priced loan, the selected input features (LTV ratio, DTI ratio, property value, loan amount, introductory interest rate, customer income, etc.) are likely to be some of the most influential factors that a lender would consider. See Appendix \ref{a_sec:mort} for general information regarding HMDA data.

@@

\noindent As in unconstrained GBM, $\Theta^{\text{mono}}_b$ is selected in a greedy, additive fashion by minimizing a regularized loss function that considers known target labels, $\mathbf{y}$, the predictions of all subsequently trained trees in the in $g^{\text{MGBM}}$, $g^{\text{MGBM}}_{b-1}(\mathbf{X})$, and the $b$-th tree splits, $T_b(\mathbf{x}^{(i)};\Theta^\text{mono}_b)$, in a numeric error function (e.g., squared error, Huber error), $l$, and a regularization term that penalizes complexity in the current tree, $\Omega(T_b)$.

Herein, two $g^\text{MGBM}$ models are trained.  One on the simulated data and one on the mortgage data. In both cases, positive and negative monotonic constraints for each $X_j$ are selected using domain knowledge, random grid search is used to determine other hyperparameters, and five-fold cross validation and test partitions are used for model assessment.  See Appendices \ref{a_ssec:not} and \ref{a_ssec:mgbm} for details pertaining to MGBM training. For exact parameterization of the two $g^\text{MGBM}$ models, see the software resources referenced in Subsection \ref{ssec:soft}.

@@

Here, each $g^\text{XNN}$ is trained by mini-batch stochastic gradient descent (SGD) on the simulated data and mortgage data. Each $g^\text{XNN}$ is assessed in five training folds and in a test data partition. $L_1$ regularization is applied to both the projection and combination layers to induce a sparse and interpretable model, where each $n_k$ subnetwork and corresponding combination layer $\gamma_k$ are ideally associated with an important $X_j$ or combination thereof. The $g^\text{XNN}$ models appear highly sensitive to weight initialization and batch size. Be aware that $g^{XNN}$ model architectures may require manual and judicious feature selection due to long training times. For more details regarding $g^\text{XNN}$ training, see the software resources in Subsection \ref{ssec:soft} and Appendices  \ref{a_ssec:not} and \ref{a_ssec:xnn}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{One-dimensional Partial Dependence and Individual Conditional Expectation}\label{ssec:pd_ice}

PD plots are a widely-used method for describing and plotting the average predictions of a complex model $g$ across some partition of data $\mathbf{X}$ for some interesting input feature $X_j$ \cite{esl}. ICE plots are a newer method that describes the local behavior of $g$ for a single instance $\mathbf{x} \in \mathcal{X}$ \cite{ice_plots}. PD and ICE can be overlaid in the same plot to compensate for known weaknesses of PD (e.g., inaccuracy in the presence of strong interactions and correlations \cite{ice_plots}, \cite{ale_plot}), to identify interactions modeled by $g$, and to create a holistic global and local portrait of the predictions for some $g$ and $X_j$ \cite{ice_plots}. For details regarding the calculation of one-dimensional PD and ICE, see the software resources in Subsection \ref{ssec:soft} and Appendices \ref{a_ssec:not} and \ref{a_ssec:pd_ice}.

@@

Shapley explanations are a class of additive, locally accurate feature contribution measures with long-standing theoretical support \cite{shapley}, \cite{shapley1988shapley}. Shapley explanations are the only possible locally accurate and globally consistent feature contribution values, meaning that Shapley explanation values for input features always sum to $g(\mathbf{x})$ for some $\mathbf{x} \in \mathcal{X}$ and that Shapley explanation values should never decrease in magnitude for some $x_j$ when $g$ is changed such that $x_j$ truly makes a stronger contribution to $g(\mathbf{x})$ \cite{shapley}, \cite{tree_shap}.

Shapley values can be estimated in different ways, many of which are intractable for datasets with large $\mathcal{P}$. Tree SHAP is a specific implementation of Shapley explanations that relies on traversing internal decision tree structures to efficiently estimate the contribution of each $x_j$ for some $g(\mathbf{x})$ \cite{tree_shap}. Tree SHAP (SHapley Additive exPlanations) has been implemented efficiently in popular gradient boosting libraries such as \href{http://docs.h2o.ai/h2o/latest-stable/h2o-py/docs/modeling.html#h2ogradientboostingestimator}{\texttt{h2o}}, \href{https://lightgbm.readthedocs.io/en/latest/Python-Intro.html}{\texttt{LightGBM}}, and \href{https://xgboost.readthedocs.io/en/latest/python/python_intro.html}{\texttt{XGBoost}}, and Tree SHAP is used to calculate accurate and consistent global and local feature importance for MGBM models in Sections \ref{sec:m_and_m} and \ref{sec:res}. Deep SHAP is an approximate Shapley value technique that creates SHAP values for ANNs \cite{shapley}. Deep SHAP is implemented in the \href{https://github.com/slundberg/shap}{\texttt{shap}} package and is used to generate SHAP values for the two $g^{XNN}$ models discussed in Sections \ref{sec:m_and_m} and \ref{sec:res}. For more information pertaining to the calculation of Shapley values, see Appendices \ref{a_ssec:not} and \ref{a_ssec:shap}.

@@

Because many technical and academic discussions of fairness in ML have been inconclusive\footnote{See: \href{https://www.youtube.com/watch?v=jIXIuYdnyyk}{Tutorial: 21 Fairness Definitions and Their Politics}.}, this text will draw on regulatory and legal standards that have been used for years in industries like Financial Services. The discussed metrics are also representative of fair lending analyses and pair well with the mortgage data. See Appendix \ref{a_sec:types} for a brief discussion regarding different types of discrimination under US laws and Appendix \ref {a_sec:prac_v_stat} for remarks on statistical vs. practical significance for discrimination measure values.

One measure is known as marginal effects (ME). ME is simply the difference between the percent of the control group members receiving a favorable outcome and the percent of the protected class members receiving a favorable outcome.

@@

Tables \ref{tab:sim_n_acc} and \ref{tab:sim_disc} show the results of the disparity tests using the simulated data for two hypothetical sets of class-control groups.  Several measures of disparities are shown, with the SMDs calculated using the probabilities from $g^{\text{MGBM}}$ and  $g^{\text{XNN}}$, and the false positive rates (FPRs), false negative rates (FNRs), their ratios, MEs, and AIRs calculated using a binary outcome based on a cutoff of 0.6 (anyone with probabilities of 0.6 or greater receives the favorable outcome).\footnote{See Appendix \ref{a_sec:cut} for comments pertaining to cutoff selection and discrimination testing.}

Since $g^{\text{MGBM}}$ and $g^{\text{XNN}}$ assume that a higher score is favorable (as might be the case if the model were predicting responses to marketing offers), one might consider the relative FNRs as a measure of the class-control disparities. Table \ref{tab:sim_disc} shows that protected group 1 has higher relative FNRs under $g^{\text{XNN}}$ (1.13 vs. 1.06).  However, in Table \ref{tab:sim_n_acc} the overall FNRs were lower for  $g^{\text{XNN}}$ (0.357 vs. 0.401). This illustrates a danger in considering relative class-control metrics in isolation when comparing across models: despite the $g^\text{MGBM}$ appearing to be a relatively fairer model, more protected group 1 members experience negative outcomes using $g^{\text{MGBM}}$. This is because FNR accuracy improves for both the protected group 1 and control group 1, but members of control group 1 benefit more than those in protected group 1.  Of course, the choice of which model is truly fairer is a policy question.

\begin{table}[H]
	\captionsetup{width=7cm}
	\hskip-20pt\parbox{.48\textwidth}{
	\caption{Group size, accuracy, and FNR for $g^\text{MGBM}$ and $g^\text{XNN}$ on the simulated test data.} 
	\centering
	\begin{tabular}{ccccc}
		\toprule
		\textbf{Class}	& $\mathbf{N}$	& \textbf{Model} & \textbf{Accuracy} & \textbf{FNR} \\
		\midrule
		Protected 1	& 3,057		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.770 \\ 0.771\end{tabular} 	&  \begin{tabular}{@{}c@{}} 0.401 \\ 0.357\end{tabular} \\	
		\midrule			 
		Control 1		& 16,943		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.739 \\ 0.756\end{tabular}	& \begin{tabular}{@{}c@{}} 0.378 \\ 0.314\end{tabular} \\
		\midrule
		Protected 2	& 9,916		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.758 \\ 0.762\end{tabular}	&  \begin{tabular}{@{}c@{}} 0.331 \\ 0.302\end{tabular} \\	
		\midrule
		Control 2		& 10,084		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.729 \\ 0.756\end{tabular}	& \begin{tabular}{@{}c@{}} 0.420 \\ 0.332\end{tabular} \\		
		\bottomrule
		\label{tab:sim_n_acc}
	\end{tabular}
	}\hspace{10pt}
	\parbox{.48\textwidth}{
	\vspace{-13pt}
	\caption{AIR, ME, SMD, and FNR ratio for $g^\text{MGBM}$ and $g^\text{XNN}$ on the simulated test data.}
	\centering
	\begin{tabular}{ccccccc}
		\toprule
		\textbf{Model}		&  \begin{tabular}[x]{@{}c@{}}\textbf{Protected}\\\textbf{Class}\end{tabular} & \begin{tabular}[x]{@{}c@{}}\textbf{Control}\\\textbf{Class}\end{tabular}					& \textbf{AIR}									& \textbf{ME}									& \textbf{SMD} &  \begin{tabular}[x]{@{}c@{}}\textbf{FNR}\\\textbf{Ratio}\end{tabular}\\
		\midrule
		$g^\text{MGBM}$	& \begin{tabular}{@{}c@{}} 1 \\  2\end{tabular}		& \begin{tabular}{@{}c@{}} 1 \\ 2\end{tabular}	& \begin{tabular}{@{}c@{}} 0.752 \\ 1.10\end{tabular}	& \begin{tabular}{@{}c@{}} 9.7\% \\ -3.6\%\end{tabular}	& \begin{tabular}{@{}c@{}} -0.206 \\ 0.106\end{tabular}	& \begin{tabular}{@{}c@{}} 1.06 \\ 0.788\end{tabular}\\
		\midrule 
		$g^\text{XNN}$		& \begin{tabular}{@{}c@{}} 1 \\ 2\end{tabular}		& \begin{tabular}{@{}c@{}} 1 \\  2\end{tabular}	& \begin{tabular}{@{}c@{}} 0.727 \\ 0.976\end{tabular}	& \begin{tabular}{@{}c@{}} 12.0\%\\ 1.0\%\end{tabular}	& \begin{tabular}{@{}c@{}} -0.274 \\ 0.001\end{tabular}	& \begin{tabular}{@{}c@{}} 1.13 \\ 0.907\end{tabular} \\										
		\bottomrule
		\label{tab:sim_disc}
	\end{tabular}
	}
\end{table}

@@

@@

\begin{table}[H]
	\captionsetup{width=7cm}
	\hskip-20pt\parbox{.48\textwidth}{
		\caption{Group size, accuracy, and FNR for $g^\text{MGBM}$ and $g^\text{XNN}$ on the mortgage test data.} 
		\centering
		\begin{tabular}{ccccc}
\toprule
\textbf{Class}	& $\mathbf{N}$	& \textbf{Model} 												& \textbf{Accuracy} 																			& \textbf{FNR} \\
\midrule
Black 		& 2,608		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.654 \\ 0.702\end{tabular} 				& \begin{tabular}{@{}c@{}} 0.457 \\ 0.308\end{tabular} \\	
\midrule			 
White 		& 28,361		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.817 \\ 0.857\end{tabular}			& \begin{tabular}{@{}c@{}} 0.508 \\ 0.360\end{tabular}\\		
\midrule
Female 		& 8,301		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}}  0.768 \\ 0.822\end{tabular}		& \begin{tabular}{@{}c@{}} 0.402 \\ 0.322\end{tabular}\\
\midrule
Male 		& 13,166		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.785 \\	0.847\end{tabular}			& \begin{tabular}{@{}c@{}} 0.497 \\ 0.347\end{tabular}\\
\bottomrule
			\label{tab:mort_n_acc}
		\end{tabular}
	}\hspace{10pt}
	\parbox{.48\textwidth}{
		\vspace{-13pt}
		\caption{AIR, ME, SMD, and FNR ratio for $g^\text{MGBM}$ and $g^\text{XNN}$ on the mortgage test data.}
		\centering
		\begin{tabular}{ccccccc}
\toprule
\textbf{Model}		&  \begin{tabular}[x]{@{}c@{}}\textbf{Protected}\\\textbf{Class}\end{tabular} & \begin{tabular}[x]{@{}c@{}}\textbf{Control}\\\textbf{Class}\end{tabular}					& \textbf{AIR}									& \textbf{ME}									& \textbf{SMD} &  \begin{tabular}[x]{@{}c@{}}\textbf{FNR}\\\textbf{Ratio}\end{tabular}\\
\midrule
$g^\text{MGBM}$	& \begin{tabular}{@{}c@{}} Black \\ Female\end{tabular}		& \begin{tabular}{@{}c@{}} White \\ Male\end{tabular}	& \begin{tabular}{@{}c@{}} 0.776 \\ 0.948\end{tabular}	& \begin{tabular}{@{}c@{}} 18.3\% \\ 4.1\%\end{tabular}	& \begin{tabular}{@{}c@{}} 0.628 \\ 0.084\end{tabular}	& \begin{tabular}{@{}c@{}} 0.900 \\ 0.810\end{tabular} \\
\midrule 
$g^\text{XNN}$		& \begin{tabular}{@{}c@{}} Black \\ Female\end{tabular}		& \begin{tabular}{@{}c@{}} White \\ Male\end{tabular}	& \begin{tabular}{@{}c@{}} 0.743 \\ 0.955\end{tabular}	& \begin{tabular}{@{}c@{}} 21.4\% \\ 3.6\%\end{tabular}	& \begin{tabular}{@{}c@{}} 0.621 \\ 0.105\end{tabular}	& \begin{tabular}{@{}c@{}} 0.855 \\ 0.927\end{tabular} \\							
\bottomrule
			\label{tab:mort_disc}
		\end{tabular}
	}
\end{table}

@@

@@

The heightened interest in exploring and fixing algorithmic bias has led to a significant body of work focused in using ML to diminish discrimination in model outcomes \cite{friedler2019comparative}. Broadly, these can be defined into two groups: more traditional methods that mitigate discrimination by searching across possible algorithmic and feature specifications, and more recently developed approaches that change the algorithms or input data themselves in order to mitigate disparities. Many approaches that have been developed in the last five to seven years focus on altering the algorithm itself, preprocessing the data, or post-processing the predictions in order to diminish class-control correlations or dependencies. For comments on why these techniques could result in regulatory non-compliance in certain cases, see Appendix \ref{a_sec:comp}.
